---
title: "BIOL4800/6220_Fall2022_RTutorial"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

# R Basic

## Introduction

### R as a calculator

```{r}
1 + 100
```

```{r}
2^2
```

```{r}
7/2
```

```{r}
4-1
```

```{r}
2*2
```

```{r}
3+5*2
```

```{r}
(3+5)*2
```

```{r}
2e2
```

```{r}
2/10000
```

```{r}
log(1)
```

```{r}
exp(0)
```

```{r}
1 == 1
```

```{r}
1 >= 1
```

```{r}
1 > 1
```

```{r}
1 != 2
```

```{r}
x <- 10
x
```

```{r}
x + x
```

```{r}
x - 5
```

```{r}
x^2
```

```{r}
y <- x+x
y
```

### Vectorization

```{r}
1:5
```

```{r}
2^(1:5)
```

```{r}
v <- 1:5
2^v
```

```{r}
log10(v)
```

### Environment

```{r}
ls()
```

```{r}
#rm(v)
```


```{r}
ls()
```

## Package Management

```{r}
installed.packages()
```


```{r}
#install.packages("vegan", dependencies=TRUE)
```

```{r}
#library(vegan)
```



```{r}
#remove.packages("vegan")
```

## Project Management

1. Set up management structure (i.e., "data", "src", "results", "doc")
2. Discuss avoiding redundancy of files
3. Make sure to use version control(i.e, Git)

## Getting Help
```{r}
?plot
```

```{r}
help(plot)
```

```{r}
?"<-"
```

```{r}
#vignette("FAQ-vegan")
```

```{r}
#citation("vegan")
```

## Data and their formats

```{r}
cats <- data.frame(coat = c("calico","black","tabby"),
                   weight = c(2.1,5,3.2),
                   likes_string = c(1,0,1))
```


```{r}
cats
```

```{r}
getwd()
```

```{r}
write.csv(cats, file = "./data/cats.csv")
```

```{r}
cats <- read.csv("./data/cats.csv")
```

```{r}
cats$coat
```

```{r}
cats$weight
```

```{r}
cats$weight*10
```

```{r}
log(cats$weight)
```

```{r}
logweight <- log(cats$weight)
cbind(cats,logweight)
```

```{r}
cats <- cbind(cats,logweight)
```

```{r}
cats
```

```{r}
paste("My cat is ", cats$coat, ", and it weighs ", cats$weight, "kg, ", sep="")
```

### Data Types

```{r}
typeof(cats$coat)
```

```{r}
typeof(cats$weight)
```

```{r}
typeof(cats$likes_string)
```

```{r}
typeof(3.14159)
```

```{r}
typeof(1i)
```

```{r}
typeof(FALSE)
```

```{r}
typeof(TRUE)
```

```{r}
typeof(T)
typeof(F)
```

```{r}
typeof(cats)
```

```{r}
class(cats)
```

```{r}
file.show("./data/cats.csv")
```

```{r}
cats$likes_string <- as.logical(cats$likes_string)
typeof(cats$likes_string)
```

```{r}
cats$likes_string
```

```{r}
c("a", "b")
```

```{r}
ab <- c("a", "b")
ab
```

```{r}
c(ab, "c")
```

```{r}
c(ab, ab)
```

```{r}
1:10
```

```{r}
seq(10)
```

```{r}
z <- seq(10)
head(z, n=3)
```

```{r}
length(z)
```

```{r}
class(z)
```

```{r}
typeof(z)
```

```{r}
seq(0,100, by=5)
```

### Data Frames

```{r}
cats
```

```{r}
str(cats$coat)
```

```{r}
coats <- c("tabby", "tortoiseshell", "tortoiseshell", "black", "tabby")
coats
```

```{r}
str(coats)
```

```{r}
factor(coats)
```

```{r}
categories <- factor(coats)
class(coats)
class(categories)
```

```{r}
str(categories)
```

### Lists

```{r}
list_example <- list(title="Numbers", numbers = 1:10, data=TRUE)
```

```{r}
list_example
```

```{r}
another_list <- list(1,"a",TRUE,1+1i)
another_list
```

```{r}
typeof(list_example)
class(list_example)
```

```{r}
typeof(cats)
class(cats)
```

```{r}
data.frame(list_example)
```

```{r}
cats
```

```{r}
cats[,3]
```

```{r}
cats[3,]
```

```{r}
cats[2:3,c(1,3)]
```

### Matricies

```{r}
matrix_example <- matrix(0, ncol=5, nrow=3)
```

```{r}
matrix_example
```

```{r}
class(matrix_example)
```

```{r}
typeof(matrix_example)
```

```{r}
str(matrix_example)
```

```{r}
dim(matrix_example)
```

```{r}
ncol(matrix_example)
```

```{r}
nrow(matrix_example)
```

```{r}
class(data.frame(matrix_example))
```

```{r}
df_example <- data.frame(matrix_example)
```

```{r}
df_example
```

### Subsetting

```{r}
p <- c(2.3, 6.9, 4.0, 23, 1)
p
```

```{r}
names(p) <- c('a', 'b', 'c', 'd', 'e')
p
```

```{r}
p[1]
```

```{r}
p[2:4]
```

```{r}
p[c(1,5)]
```

```{r}
p[c(1,1,1,3,5,5)]
```

```{r}
p[6]
```

```{r}
p[-3]
```

```{r}
p[c(-1,-5)]
```

```{r}
p[-(2:4)]
```

```{r}
p[c('a','c')]
```

```{r}
p[c(T,F,T,F,T)]
```

```{r}
p[names(p) != 'c']
```

### Matricies Resumed

```{r}
set.seed(65)
m <- matrix(rnorm(6*4), ncol=4, nrow=6)
m
```

```{r}
m[3:4, c(3,1)]
```

```{r}
m[,c(3,4)]
```

### Lists Revisited

```{r}
xlist <- list(a="BIOL48006220",b=seq(1,10,by = 0.5), data="Grade")
xlist
```

```{r}
xlist[1:2]
```

```{r}
xlist['a']
```

```{r}
xlist[['a']]
```

```{r}
xlist['b']
```

```{r}
xlist[['b']]
```

```{r}
xlist$b
```

### Data Frames Revisited

```{r}
gp <- read.csv('./data/forest_area_sq_km.csv')
```

```{r}
head(gp, n=10L)
```

```{r}
head(gp[3], n=10L)
```

```{r}
nrow(gp)
```

```{r}
head(gp[["country"]], 10L)
```

```{r}
gp$X2001
```

```{r}
gp[1:3, 2:5]
```

```{r}
gp[ which(gp$X1989 <= 100 & gp$X2019 >= 100),]
```

## Conditionals and Flow

```{r}
n <- 10

if (n < 10){
  print("n is less than 10")
} else if (n > 10){
  print("n is greater than 10")
}else{
  print("n is equal to 10")
}
```

```{r}
for (n in seq(0,20)){
  if (n < 10){
    print("n is less than 10")
}   else if (n > 10){
    print("n is greater than 10")
}   else{
    print("n is equal to 10")
}
  
}
```

```{r}
g <- 0

while(g <= 10){
  print(paste(g,"is less than or equal to 10"))
  g <- g+1
}
```

## Plotting and Figures

```{r}
#install.packages("ggplot2")
```

```{r}
library(ggplot2)
```

```{r}
gp <- read.csv('./data/gapminder_all.csv')

head(gp)
```


```{r}
ggplot(data=gp, mapping=aes(x=gdpPercap_1952, y=pop_1952)) + geom_point()
```

```{r}
ggplot(data=gp, mapping=aes(x=gdpPercap_2002, y=pop_2002)) + geom_point()
```

```{r}
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv")
head(gapminder)
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country, color=continent)) + geom_line()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country)) + geom_line(mapping=aes(color=continent)) + geom_point()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(alpha=0.25) + scale_x_log10()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(alpha=0.25, color="purple", size = 0.9) + scale_x_log10() + geom_smooth(method=lm, color="gold", size = 1.25)
```

```{r}
africas <- gapminder[gapminder$continent == "Africa", ]
head(africas)
```

```{r}
ggplot(data=africas, mapping=aes(x=year, y=lifeExp)) + 
  geom_line(color = "red") + 
  facet_wrap( ~ country) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(
    x = "Year",
    y = "Life Expectancy",
    title = "Life Expectancy Over Time in African Countries",
  )
```

```{r}
AfricanLifeExp <- ggplot(data=africas, mapping=aes(x=year, y=lifeExp)) + 
  geom_line(color = "red") + 
  facet_wrap( ~ country) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(
    x = "Year",
    y = "Life Expectancy",
    title = "Life Expectancy Over Time in African Countries",
  )
```

```{r}
ggsave(filename = "data/AfricanLifeExp.png", plot = AfricanLifeExp, width = 24, height = 40, dpi = 300, units = "cm")
```

```{r}
pdf(file = "results/AfricanLifeExp.pdf", width = 24, height = 40)
plot(AfricanLifeExp)
dev.off()
```

```{r}
write.table(gapminder, file = "data/gapminder_web.csv", sep = ",")
```

```{r}
write.csv(africas, file="data/gapminder_web_africas.csv")
```

### Fancy Plots

```{r}
#install.packages(c("ggridges","viridis","hrbrthemes"), dependencies = T)
```

```{r}
spider_data <- read.csv(file="https://wsc.nmbe.ch/resources/species_export_20221025.csv", header=TRUE)
spider_data
```




```{r}
install.packages(c("ggstatsplot","palmerpenguins","tidyverse"), dependencies=T)
library(ggstatsplot)
library(palmerpenguins)
```


## Statistics

### Addition ways of importing/reading and manipulating data

```{r}
rand <- c(12, 54, 98, 65, 38)
rand
```

```{r}
sum(rand)
```

```{r}
length(rand)
```

```{r}
avg.rand <- sum(rand)/length(rand)
avg.rand
```

```{r}
mean(rand)
```

```{r}
sort(rand)
```

```{r}
min(rand)
max(rand)
```

```{r}
cumsum(rand)
```

```{r}
diff(rand)
```

```{r}
rand[2]
```

```{r}
rand*2
```

### Analyses

#### Univariate Statistics

1. Categorical Data
  a. Barplots
```{r}
beer <- c(3, 4, 1, 1, 3, 4, 3, 3, 1, 3, 2, 1, 2, 3, 2, 3, 1, 1, 1, 1, 4, 3, 1)
length(beer)
```

```{r}
barplot(beer)
```

```{r}
table(beer)
```

```{r}
barplot(table(beer), xlab ="Beer", ylab="Frequency") # Counts
```

```{r}
barplot(table(beer)/length(beer), xlab="Beer", ylab="Proportion") # Proportions instead of Frequencys
```

  b. Pie Charts
```{r}
pie(table(beer), main="beer preference by students")
```

2. Numerical Data
  a. Stem-and-leaf Plots
```{r}
stem(beer)
```

  b. Strip Chart
```{r}
stripchart(beer, method="stack")
```

3. Measures of Center
  a. Mean
```{r}
mean(beer)
```
```{r}
mean(gapminder$lifeExp)
```



  b. Median
```{r}
median(beer)
```

```{r}
median(gapminder$lifeExp)
```

```{r}
with(gapminder, median(lifeExp))
```

  c. Mode
```{r}
which(table(beer) == max(table(beer)))
```

4. Variation
  a. Range
```{r}
range(beer)
```

```{r}
diff(range(beer))
```

  b. Variants
```{r}
var(beer)
```

```{r}
sd(beer)
```
  c. IQRs
```{r}
IQR(beer)
```

  d. Z-scores
```{r}
scale(beer)
```

  e. Summary
```{r}
summary(beer)
```

5. Plots
  a. Histograms
```{r}
hist(beer, breaks="scott")
```

```{r}
hist(beer, breaks = "scott", prob=TRUE)
```

```{r}
hist(beer, breaks = "scott", prob=T)
lines(density(beer))
```

```{r}
plot(density(beer))
```

  b. Box Plots
```{r}
boxplot(beer)
```


```{r}
summary(beer)
```

#### Bivariate and Multivariate Statistics
1. Plotting and Regression
  a. Box Plotting

```{r}
spid.gen <- read.csv("data/spider_genitalia.csv", header = T)
spid.gen
```

```{r}
boxplot(spid.gen$left.bulb/spid.gen$carapace.length ~ spid.gen$habitat)
```

  b. Scatter Plotting
 
```{r}
plot(spid.gen$left.bulb ~ spid.gen$right.bulb)
```

```{r}
plot(spid.gen$left.bulb ~ spid.gen$right.bulb, pch = as.character(spid.gen$habitat))
```

  c. Linear Regression
  
```{r}
gen.reg <- lm(spid.gen$left.bulb ~ spid.gen$right.bulb)
gen.reg
```

```{r}
summary(gen.reg)
```

```{r}
plot(spid.gen$left.bulb ~ spid.gen$right.bulb, pch = as.character(spid.gen$habitat))
abline(gen.reg)
```

  d. Correlation Coefficeints & Spearman Rank Correlation
  
```{r}
cor.gen <- with(spid.gen, cor(left.bulb, right.bulb))
cor.gen^2
```

```{r}
spearman.cor.gen <- with(spid.gen, cor(left.bulb, right.bulb, method ="spearman"))
spearman.cor.gen^2
```

  e. Residuals
  
```{r}
residuals(gen.reg)
```

  f. Transformations
  
```{r}
plot(log10(spid.gen$left.bulb) ~ spid.gen$right.bulb)
```

2. Comparing Discrete Treatment Effects
  a. Chi-squared
  
```{r}
obs_weighted <- c(4,15,6,15,18,2)
obs_fair <- c(10,10,10,10,10,10)
exp <- c(.16,.17,.16,.17,.17,.17)
chisq.test(obs_weighted, p=exp)
chisq.test(obs_fair, p=exp)
```

  b. T-tests

```{r}
#HO - true mean is equal to 0, but you can set true mean to other values with mu=
t.test(spid.gen$carapace.length)

t.test(spid.gen$carapace.length, mu=29)
```

```{r}
habitat.t2 <- t.test(spid.gen$left.bulb ~ spid.gen$habitat)
habitat.t2
```

```{r}
habitat.tless <- t.test(spid.gen$left.bulb, alternative = "less")
habitat.tless
```

```{r}
habitat.tgreater <- t.test(spid.gen$left.bulb, alternative = "greater")
habitat.tgreater
```





